Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2014-05-07 manuscripts/manuscript.Rmd
support for linenumbering, fix refs, doublespace, add refs, add tex output

---
title: "Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if the exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points leading to harvest policies
          that guarantee extinction. The Gaussian process based approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches.

@@

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated.
This extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
alogrithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013].

@@

<!-- necessary? --> <!-- should reflect the dual problem of extrapolation and model uncertainty better -->

This paper represents the first time the SDP decision-making framework has
been used without an a priori model of the underlying dynamics.  In contrast to parametric models which
can only reflect uncertainty in parameter estimates, the BNP approach
provides a more state-space dependent representation of uncertainty.
This permits a much greater uncertainty far from the observed data than
near the observed data.  These features allow the GPDP approach to
find robust management solutions in face of limited data and without
knowledge of the correct model structure.

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@Sugeno2013], estimate the maximum reproductive
rate [@Sugeno2013], determine temporal variation in food
availability [@Sigourney], and provide a basis for identifying
model-mispecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric monicker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions  [e.g. @Lehmann1975]. 
However, we retain the terminology for consistency with previous
studies.

@@

We infered posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results wer robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

Using the simulated training data we also estimate a Gaussian Process defined by a radial basis
function kernel of two parameters: $\ell$, which gives the characteristic
length-scale over which correlation between two points in state-space
decays, and $\sigma$, which gives the scale of the process noise by which
observations $Y_{t+1}$ may differ from their predicted values $X_{t+1}$
given an observation of the previous state, $X_t$. @Munch2005a gives an
accessible introduction to the use of Gaussian Processes in providing a
Bayesian nonparametric description of the stock-recruitment relationship.

@@

It is reasonably straight-forward to derive the harvest policy from the stimated BP by inserting it into a Stochastic Dynamic
Programming (SDP) algorithm.  Recall that unlike the parametric models the Gaussian
process with fixed parameters already predicts a distribution of
curves rather than a single curve. We must first integrate over 
this distribution of curves given a sampling of parameter values drawn
from the posterior distribution of the two GP parameters, before
integrating over the posterior of those parameters themselves.

@@

@@

All models fit the observed data rather closely and with relatively small uncertainty, as illustrated in the posterior predictive curves in Figure 1.  Figure 1 shows the training data of stock sizes observed over time as points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions.  Each model manages to fit the observed data rather closely. Compared to the true model most estimates appear to overfit, predicting fluctuations that are actually due purely to stochasticity.  All model selection criteria shown in Table 1 penalize more complex models and show a preference for the simpler Ricker model over the more complicated alternate stable state models (Allen and Myers).  Details on MCMC estimates for each model, traces, and posterior distributions can be found in the appendix.

@@

```{r out_of_sample_predictions, include=TRUE, echo=FALSE, fig.width=8, fig.height=6, fig.cap="Out of sample predictions of the dynamics under each model.  Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always predicts population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that the GP is both more pessimistic and more uncertain about the future state than the paramteric models, while the realized state often falls outside of the expected range forecast by the structurally incorrect Myers and Ricker models." }
```

@@

These results are robust across a range of stochastic realizations,
models, and parameter values.  The results across this range can most
easily be compared by using the relative differences in net present
value realized by each of the model. Figure 6 shows the distribution of values
(profits) realized by each simulation shown in Figure 5.  The GPDP
approach most consistently realizes a value close to the optimal solution,
and importantly avoids ever driving the system across the tipping point,
which results in the near-zero value cases in the parametric models.

@@

These results are not sensitive to the modeling details of the simulation.
The GPDP estimate remains very close to the optimal solution obtained
by knowning the true model across changes to the training simulation,
noise scale, parameters or structure of the underlying model, as seen in
Figure 7. The sensitivity analysis uses a factorial design of eight replicates 
of three different randomly generated parameter sets for each of two different 
generating models (Allen and Myers) over two different noise levels 
(0.01 and 0.05), for a total of 8 x 3 x 2 x 2 = 96 scenarios. In Figure 7, 
results are pooled across different random seeds (111, 222,
333), noise values of 0.01 and 0.05 (not large enough to violate the
self-sustaining criterion of @Reed1979), and different randomly generated
parameters sets for each model, and using either the Myers or Allen as the
underlying structure.  While the GPDP performs slightly closer to optimal
on the smaller noise levels, and closer on the Allen model then the Myers
model, the differences are small compared to the values observed in using
the wrong parametric model.  The GPDP solution consistently avoids either
failing to harvest at all or crashing the stock, which results in the very
low profits/values seen in parametric models in the previous figure. The
sensitivity analysis is compared against the optimal solution alone and
not shown against the estimates of the various parameteric models (but
see Figure 6), where convergence of the Gibbs sampling could contribute
to their poorer performance if not checked by hand for each configuration.

```{r sensitivity, fig.height=6, dependson="sensitivity-calc", echo=FALSE, fig.cap="Sensitivity Analysis.  Histogram shows the ratio of the realized net present value / profits derived when managing under the GPDP Gaussian Process approach over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels."}
ggplot(dat) + geom_histogram(aes(value, fill=noise)) + xlim(0,1.0) + theme_bw() + xlab("value as fraction of the optimal") + facet_wrap(~model, ncol=1)
```

@@

In any modeling effort, models must be chosen for the task at hand.
Though simple mechanistically motivated models offer the greatest
potential to increase our basic understanding of ecological processes
[@Cuddington2013; @Geritz2012], such models can be not only inaccurate but
misleading when relied upon in a quantitative decision making framework.
In this paper we have tackled two aspects of uncertainty that are both
common to many ecological desicion-making problems and fundamentally
challenging to existing approaches which largely rely on parametric
models:

@@

@@

$$x|y \sim \mathcal{N}(E,C)$$
$$E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y$$
$$C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)$$

@@

@@

The Gaussian process priors on both the lengthscale $\ell$ and process noise $\sigma$ are are inverse Gamma distributed,

@@

Form this we can write down the Bellman recursion as:

@@

Training data
-------------

Eacho of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = $ `r p[1]`, 
$K =$ `r p[2]`, $C =$ `r p[3]`, and  $\sigma_g =$ `r sigma_g` 
for $T=$ `r Tobs` timesteps, starting at initial condition $X_0 = $ `r Xo`. 
The training data can be seen in Figure 1.

@@

A copy of the script to reconstruct the simulations and analysis shown here is provided in the supplemental materials, and through this version-stable link to the project's Github code repostitory, [nonparameteric-bayes.R](https://github.com/cboettig/nonparametric-bayes/blob/371114ec43bb32e92f330f049b420047edae5eb0/inst/doc/nonparametric-bayes.R).  This code is dynamically embedded into the manuscript using Knitr, [@knitr].  The script relies on custom routines for executing the estimation of the Gaussian process and the for solving the stochastic dynamic programming problem.  The sensitivity analysis is provided as a separate script, [sensitivity.Rmd](https://github.com/cboettig/nonparametric-bayes/blob/62e0f81f9e8eabd0f7ee462f098320409a407f81/inst/examples/sensitivity.Rmd). These routines are provided as an R package, [nonparameteric-bayes](https://github.com/cboettig/nonparametric-bayes), also available on Github.

Data
----

>>>>

---
layout: 12pt,review
title: "Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if the exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The Gaussian process approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches.

@@

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated.
This extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013].

@@

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics.  In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a more
state-space dependent representation of uncertainty.  This permits
a much greater uncertainty far from the observed data than near the
observed data.  These features allow the GPDP approach to find robust
management solutions in face of limited data and without knowledge of
the correct model structure.

@@

We use a Bayesian nonparameteric regression approach based on Gaussian process priors. 
The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
GP models have subsequently been used to test for the presence of Allee
effects [@Sugeno2013], estimate the maximum reproductive
rate [@Sugeno2013], determine temporal variation in food
availability [@Sigourney2012], and provide a basis for identifying
model-misspecification [@Thorson2014].
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006. The nonparametric moniker for
the GP approach is unfortunate as (a) it still involves the estimation
of parameters, and (b), statisticians also use non-parametric to mean
models that make only weak distributional assumptions  [e.g. @Lehmann1975]. 
However, we retain the terminology for consistency with previous
studies.

@@

We inferred posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results were robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

Using the simulated training data we also estimate a Gaussian process defined by a radial basis
function kernel of two parameters: $\ell$, which gives the characteristic
length-scale over which correlation between two points in state-space
decays, and $\sigma$, which gives the scale of the process noise by which
observations $Y_{t+1}$ may differ from their predicted values $X_{t+1}$
given an observation of the previous state, $X_t$. @Munch2005a gives an
accessible introduction to the use of Gaussian processes in providing a
Bayesian nonparametric description of the stock-recruitment relationship.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a Stochastic Dynamic Programming (SDP)
algorithm.  Recall that unlike the parametric models the Gaussian process
with fixed parameters already predicts a distribution of curves rather
than a single curve. We must first integrate over this distribution of
curves given a sampling of parameter values drawn from the posterior
distribution of the two GP parameters, before integrating over the
posterior of those parameters themselves.

@@

To ensure our results are robust to our choice of parameters, 
we perform our analysis over 96 different scenarios. To better
understand the process, we will first describe in detail the
results of a single scenario.

@@

All models fit the observed data rather closely and with relatively small uncertainty, as illustrated in the posterior predictive curves in Figure 1.  Figure 1 shows the training data of stock sizes observed over time as points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions.  Each model manages to fit the observed data rather closely. Compared to the true model most estimates appear to over-fit, predicting fluctuations that are actually due purely to stochasticity.  All model selection criteria shown in Table 1 penalize more complex models and show a preference for the simpler Ricker model over the more complicated alternate stable state models (Allen and Myers).  Details on MCMC estimates for each model, traces, and posterior distributions can be found in the appendix.

@@

```{r out_of_sample_predictions, include=TRUE, echo=FALSE, fig.width=8, fig.height=6, fig.cap="Out of sample predictions of the dynamics under each model.  Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always predicts population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that the GP is both more pessimistic and more uncertain about the future state than the parametric models, while the realized state often falls outside of the expected range forecast by the structurally incorrect Myers and Ricker models." }
```

@@

The results across replicate stochastic simulations can most
easily be compared by using the relative differences in net present
value realized by each of the model. Figure 6 shows the distribution of values
(profits) realized by each simulation shown in Figure 5.  The GPDP
approach most consistently realizes a value close to the optimal solution,
and importantly avoids ever driving the system across the tipping point,
which results in the near-zero value cases in the parametric models.

@@

These results are not sensitive to the modeling details of the simulation.
The GPDP estimate remains very close to the optimal solution obtained
by knowing the true model across changes to the training simulation,
noise scale, parameters or structure of the underlying model, as seen in
Figure 7. The sensitivity analysis uses a factorial design of eight replicates 
of three different randomly generated parameter sets for each of two different 
generating models (Allen and Myers) over two different noise levels 
(0.01 and 0.05), for a total of 8 x 3 x 2 x 2 = 96 scenarios. In Figure 7, 
results are pooled across different random seeds (111, 222,
333), noise values of 0.01 and 0.05 (not large enough to violate the
self-sustaining criterion of @Reed1979), and different randomly generated
parameters sets for each model, and using either the Myers or Allen as the
underlying structure.  While the GPDP performs slightly closer to optimal
on the smaller noise levels, and closer on the Allen model then the Myers
model, the differences are small compared to the values observed in using
the wrong parametric model.  The GPDP solution consistently avoids either
failing to harvest at all or crashing the stock, which results in the very
low profits/values seen in parametric models in the previous figure. The
sensitivity analysis is compared against the optimal solution alone and
not shown against the estimates of the various parametric models (but
see Figure 6), where convergence of the Gibbs sampling could contribute
to their poorer performance if not checked by hand for each configuration.

```{r sensitivity, fig.height=6, dependson="sensitivity-calc", echo=FALSE, fig.cap="Sensitivity Analysis.  Histogram shows the ratio of the realized net present value / profits derived when managing under the GPDP Gaussian process approach over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels."}
ggplot(dat) + geom_histogram(aes(value, fill=noise)) + xlim(0,1.0) + theme_bw() + xlab("value as fraction of the optimal") + facet_wrap(~model, ncol=1)
```

@@

In any modeling effort, models must be chosen for the task at hand.
Though simple mechanistically motivated models offer the greatest
potential to increase our basic understanding of ecological processes
[@Cuddington2013; @Geritz2012], such models can be not only inaccurate but
misleading when relied upon in a quantitative decision making framework.
In this paper we have tackled two aspects of uncertainty that are both
common to many ecological decision-making problems and fundamentally
challenging to existing approaches which largely rely on parametric
models:

@@

Gaussian process model
----------------------

@@

$$f(x) \sim \mathcal{N}(E,C)$$
$$E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y$$
$$C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)$$

@@

Where $K$ is the Gaussian process kernel.  Throughout, we use the radial basis function kernel:

$$ K(x,y) = \exp\left(\frac{-(x-y)^2}{2 \ell^2} \right)$$

@@

The Gaussian process priors on both the lengthscale $\ell$ and process noise $\sigma$ parameters are both vague inverse Gamma,

@@

From this we can write down the Bellman recursion as:

@@

Training data (Figures 1-6)
----------------------------

Each of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = $ `r p[1]`, 
$K =$ `r p[2]`, $C =$ `r p[3]`, and  $\sigma_g =$ `r sigma_g` 
for $T=$ `r Tobs` timesteps, starting at initial condition $X_0 = $ `r Xo`. 
The training data can be seen in Figure 1.

Training data for sensitivity analyses
-------------------------------------

A further 96 unique training data sets are generated for the sensitivity analysis, as described in the main text.

@@

All code used in producing this analysis can be found in the R package accompanying the paper, available from [http://github.com/cboettig/nonparametric-bayes]. The code for all results shown here is dynamically embedded into the manuscript using Knitr, [@knitr].

<!--
Data
----
